---
title: "Code for Carrera et al. AJHB 2026"
author: "Sofia Carrera"
date: "2025-09-15"
output: html_document
editor_options: 
  chunk_output_type: console
---


### Load packages
```{r lib}
library(tidyverse)
library(MASS)
library(glmmTMB)
library(DHARMa)
library(discSurv)
library(survival)
library(survminer)
```

Function for transparent colors
```{r make transparent}
makeTransparent <- function(black, alpha = 150){
  newColor <- col2rgb(black)
  apply(newColor, 2, function(curcoldata)         
  {rgb(red = curcoldata[1],
       green = curcoldata[2],
       blue = curcoldata[3],
       alpha = alpha,
       maxColorValue=  255)})
}
```

Custom ggplot theme
```{r custom theme}
### center title
custom_theme <- function() {
  theme_survminer() %+replace%
    theme(
      plot.title=element_text(hjust=0.5),
      plot.subtitle=element_text(hjust=0.5)
    )
}
```


## Load anonymized dataset

Analysis dataset (1288 IC w/ ACEs and LH data)
```{r}
ACEs <- read.csv("Anon-LH-data.csv", header = T, stringsAsFactors = F)

length(unique(ACEs$IC)) # anonymized
length(unique(ACEs$brgy)) # anonymized 

ACEs$icsex <- factor(ACEs$icsex)

table(ACEs$icsex) / nrow(ACEs)
```


## Descriptive stats 

### 1) ACEs frequency (cumulative and dimensions)

```{r prop}
### ACEs
ACEs %>%
  group_by(cumaces) %>%
  summarise(N = n(), Percent = (n() / nrow(ACEs)) * 100)

### Threat
sum(ACEs$n_threat != 0) / nrow(ACEs) 

### Deprivation
sum(ACEs$n_deprivation != 0) / nrow(ACEs) 

```


##### Fig. 1
Number of ACEs reported by sex

```{r fig 1}
## colors
tred <- makeTransparent("#CC6677", alpha = 100)
tblue <- makeTransparent("#88CCEE")

### plots 
h1 <- ACEs %>%
  group_by(icsex, ACES) %>%
  summarise(ic = n()) %>%
  group_by(icsex) %>%
  mutate(total = sum(ic), prop = ic/total) %>%
  ggplot(., aes(x = as.factor(ACES), fill = icsex)) + 
    geom_col(aes(y = prop), position = "identity") + 
    scale_fill_manual(values = c(tred, tblue), labels = c("F", "M")) +
    ylim(c(0,0.6)) +
    xlab("Cumulative ACEs") +
    ylab("Proportion") +
    labs(fill = "") +
    theme_classic() +
    theme(axis.line = element_line(linewidth = 0.35),
          axis.text = element_text(size = 11), 
          text = element_text(size = 12), 
          legend.position = c(0.9, 0.9))          

h2 <- ACEs %>%
  group_by(icsex, n_deprivation) %>%
  summarise(ic = n()) %>%
  group_by(icsex) %>%
  mutate(total = sum(ic), prop = ic/total) %>%
  ggplot(., aes(x = as.factor(n_deprivation), fill = icsex)) + 
    geom_col(aes(y = prop), position = "identity") + 
    scale_fill_manual(values = c(tred, tblue)) +
    xlab("Deprivation score") +
    ylab("") +
    theme_classic() +
    theme(axis.line = element_line(linewidth = 0.35),
          axis.text = element_text(size = 11), 
          text = element_text(size = 11), 
          legend.position = "none")

h3 <- ACEs %>%
  group_by(icsex, n_threat) %>%
  summarise(ic = n()) %>%
  group_by(icsex) %>%
  mutate(total = sum(ic), prop = ic/total) %>%
  ggplot(., aes(x = as.factor(n_threat), fill = icsex)) + 
    geom_col(aes(y = prop), position = "identity") + 
    scale_fill_manual(values = c(tred, tblue)) +
    xlab("Threat score") +
    ylab("") +
    theme_classic() +
    theme(axis.line = element_line(linewidth = 0.35),
          axis.text = element_text(size = 11), 
          text = element_text(size = 11), 
          legend.position = "none")

```

Multipanel figure
```{r fig 1}
library(cowplot)
plot2 <- plot_grid(h2, h3, nrow = 2, 
                   labels = c("b", "c"), label_size = 12, label_fontface = "plain")

fig1 <- plot_grid(h1, plot2, ncol = 2, rel_widths = c(1, 0.55), labels = c("a", ""), label_size = 12, label_fontface = "plain")

# ggsave('Figure1.png', width=12, height=9.6, units='cm', pointsize = 8, dpi = 1600)

```


### 2) Compare ACEs with prospective (adolescent) SES

##### Fig. S1
Dimensions
```{r figure s1}
tdarkred <- makeTransparent("#661100", 100)
tdgray <- makeTransparent("#888888", 100)

### deprivation (yes/no)
fig_data <- ACEs %>%
  mutate(deprivation = as.factor(case_when(
    n_deprivation == 0 ~ 0,
    T ~ 1
  )),
  threat = as.factor(case_when(
    n_threat == 0 ~ 0,
    T ~ 1
  )))

## threat
wilcox.test(x = fig_data$assets98[fig_data$threat == 1], y = fig_data$assets98[fig_data$threat == 0], alternative = "less", paired = F, conf.int = T)
summary(glm(threat ~ assets98, data = fig_data, family = "binomial"))
confint(glm(threat ~ assets98, data = fig_data, family = "binomial"))

## dep
wilcox.test(x = fig_data$assets98[fig_data$deprivation == 1], y = fig_data$assets98[fig_data$deprivation == 0], alternative = "less", paired = F, conf.int = T)
summary(glm(deprivation ~ assets98, data = fig_data, family = "binomial"))
confint(glm(deprivation ~ assets98, data = fig_data, family = "binomial"))

plot_d <- ggplot(fig_data, aes(deprivation, assets98)) +
  # geom_violin(scale = "count") +
  geom_boxplot(outliers = F, varwidth = T, notch = T) +
  geom_jitter(width = 0.2, aes(col = deprivation)) +
  scale_y_continuous(breaks = c(0,2,4,6,8,10), labels = c("0", "2", "4", "6", "8", "10")) +
  scale_x_discrete(breaks = c(0,1), labels = c("No", "Yes")) +
  scale_colour_manual(values = c(tdgray, tdarkred)) +
  xlab("Deprivation") +
  ylab("Household assets 1998") +
  annotate("text", x = 1.5, y = 10.7, label = "**", size = 5) +
  annotate("segment", x = 1, xend = 2, y = 10.6, yend = 10.6, lwd = 0.3) +
  theme_classic() +
  theme(legend.position="none", 
        axis.text = element_text(size = rel(0.8))) 


### threat (yes/no)
plot_t <- ggplot(fig_data, aes(threat, assets98)) +
  # geom_violin(scale = "count") +
  geom_boxplot(outliers = F, varwidth = T, notch = T) +
  geom_jitter(width = 0.2, aes(col = threat)) +
  scale_y_continuous(breaks = c(0,2,4,6,8,10), labels = c("0", "2", "4", "6", "8", "10")) +
  scale_x_discrete(breaks = c(0,1), labels = c("No", "Yes")) +
  scale_colour_manual(values = c(tdgray, tdarkred)) +
  xlab("Threat") +
  ylab("Household assets 1998") +
  annotate("text", x = 1.5, y = 10.7, label = " ", size = 5) +
  # annotate("segment", x = 1, xend = 2, y = 10.6, yend = 10.6, lwd = 0.3) +
  theme_classic() +
  theme(legend.position="none",
        axis.text = element_text(size = rel(0.8)))

plots_td <- list(plot_t, plot_d)

library(ggpubr)
ggarrange(plotlist = plots_td, nrow = 1, ncol = 2, labels = "auto", font.label = list(face = "plain"))

# ggsave('FigureS1.png', width=15, height=10, units='cm', pointsize = 8, dpi = 1500)

```


## Model ACEs
From baseline SES and sex

### 1) Cumulative ACEs
Range 0-7; variance > mean
```{r predict aces}
model <- ACEs

mean(model$ACES)
var(model$ACES)
var(model$ACES) / mean(model$ACES)

#### Zero-inflated neg binomial
comb_model <- glmmTMB(ACES ~ scale(assets83) + scale(meduc83) + icsex, data = model, family = nbinom2,
                      ziformula = ~ icsex)
summary(comb_model)
exp(confint(comb_model))

#### Check model
# library(DHARMa)
# simulationOutput <- simulateResiduals(fittedModel = comb_model, plot = F)
# 
# plot(simulationOutput)
# 
# testOutliers(comb_model)
# testDispersion(comb_model) 
# testZeroInflation(comb_model) 
# 
# # testing for number of 1s
# countOnes <- function(x) sum(x == 1)
# testGeneric(comb_model, alternative = "two.sided", summary = countOnes) 
# 
# testCategorical(comb_model, catPred = model$icsex)
# 
# plotResiduals(simulationOutput, form = model$assets83)
# plotResiduals(simulationOutput, form = model$meduc83)
```



### 2) Threat score
Range 0-3; variance ~ mean
```{r predict threat}
mean(model$n_threat)
var(model$n_threat)
var(model$n_threat) / mean(model$n_threat)

#### Zero-inflated Poisson
threat_model <- glmmTMB(n_threat ~ scale(assets83) + scale(meduc83) + icsex, data = model, family = poisson,
                        ziformula = ~ icsex)
summary(threat_model)
exp(confint(threat_model))

#### Check model
# simulationOutput <- simulateResiduals(fittedModel = threat_model, plot = F)
# plot(simulationOutput)
# 
# testOutliers(threat_model)
# testDispersion(threat_model) 
# testZeroInflation(threat_model) 
# testCategorical(threat_model, catPred = model$icsex)
# 
# plotResiduals(simulationOutput, form = model$assets83)
# plotResiduals(simulationOutput, form = model$meduc83)
```

### 3) Deprivation score
Range 0-3; variance ~ mean
```{r predict deprivation}
mean(model$n_deprivation)
var(model$n_deprivation)
var(model$n_deprivation) / mean(model$n_deprivation)


#### Zero-inflated Poisson
dep_model <- glmmTMB(n_deprivation ~ scale(assets83) + scale(meduc83) + icsex, data = model, family = poisson,
                        ziformula = ~ icsex)
summary(dep_model)
exp(confint(dep_model))

#### Check model
# simulationOutput <- simulateResiduals(fittedModel = dep_model, plot = F)
# plot(simulationOutput)
# 
# testOutliers(dep_model)
# testDispersion(dep_model) 
# testZeroInflation(dep_model) 
# testCategorical(dep_model, catPred = model$icsex)
# 
# plotResiduals(simulationOutput, form = model$assets83)
# plotResiduals(simulationOutput, form = model$meduc83)

```

## Compare outcomes by sex

Behavioral outcomes (remove right-censored IC)
```{r sex diff beh}
wilcox.test(ACEs$agesmoke[ACEs$smokestatus == 1 & ACEs$icsex == 1], ACEs$agesmoke[ACEs$smokestatus == 1 & ACEs$icsex == 0], conf.int = T) # p = 0.03

wilcox.test(ACEs$agedrink[ACEs$drinkstatus == 1 & ACEs$icsex == 1], ACEs$agedrink[ACEs$drinkstatus == 1 & ACEs$icsex == 0], conf.int = T) # p < 0.001

wilcox.test(ACEs$ageroman[ACEs$romstatus == 1 & ACEs$icsex == 1], ACEs$ageroman[ACEs$romstatus == 1 & ACEs$icsex == 0], conf.int = T) # p = 0.93

wilcox.test(ACEs$age1stsex[ACEs$sexstatus == 1 & ACEs$icsex == 1], ACEs$age1stsex[ACEs$sexstatus == 1 & ACEs$icsex == 0], conf.int = T) # p  < 0.001

```

##### Fig. S3
```{r fig s3}
#### BEHAVIORAL OUTCOMES (m and f)
# jpeg("FigureS3.png", width = 15, height = 15, units = "cm", res = 1500)

par(mfrow = c(2,2), xpd = T)

## first started smoking regularly (>= 1 / day)
hist(ACEs$agesmoke[ACEs$smokestatus == 1 & ACEs$icsex == 1], 
     xlim = c(5,36), ylim = c(0,252), breaks = seq(5,36,2), yaxt = "n",
     ylab = "N individuals", xlab = "", main = "First smoke", cex.lab = 1.2,
     col = tblue)
hist(ACEs$agesmoke[ACEs$smokestatus == 1 & ACEs$icsex == 0], 
     add = T, breaks = seq(5,36,2), 
     col = tred)
axis(2, las = 2)
text(x = 30, y = 100, "p < 0.05", cex = 0.9)
mtext("a", at = c(-2), cex = 1.3, line = 2)

## first started drinking
hist(ACEs$agedrink[ACEs$drinkstatus == 1 & ACEs$icsex == 1], 
     xlim = c(5,36), ylim = c(0,252), breaks = seq(5,36,2), yaxt = "n",
     ylab = "", xlab = "", main = "First drink", cex.lab = 1.2,
     col = tblue)
hist(ACEs$agedrink[ACEs$drinkstatus == 1 & ACEs$icsex == 0], 
     add = T, breaks = seq(5,36,2), 
     col = tred)
axis(2, las = 2)
legend(x = 32.5, y = 275, legend = c("F", "M"), fill = c(tred, tblue), bty = "n", cex = 1)
text(x = 30, y = 100, "p < 0.001", cex = 0.9)
mtext("b", at = c(-1), cex = 1.3, line = 2)


## first romantic relationship
hist(ACEs$ageroman[ACEs$romstatus == 1 & ACEs$icsex == 1], 
     xlim = c(5,36), ylim = c(0,252), breaks = seq(5,36,2), yaxt = "n",
     ylab = "N individuals", xlab = "Age (years)", main = "First relationship", cex.lab = 1.2,
     col = tblue)
hist(ACEs$ageroman[ACEs$romstatus == 1 & ACEs$icsex == 0], 
     add = T, breaks = seq(5,36,2), 
     col = tred)
axis(2, las = 2)
text(x = 30, y = 100, "p = 0.93", cex = 0.9)
mtext("c", at = c(-2), cex = 1.3, line = 2)


## first sexual intercourse
hist(ACEs$age1stsex[ACEs$sexstatus == 1 & ACEs$icsex == 1], 
     xlim = c(5,36), ylim = c(0,252), breaks = seq(5,36,2), yaxt = "n",
     ylab = "", xlab = "Age (years)", main = "Sexual debut", cex.lab = 1.2,
     col = tblue)
hist(ACEs$age1stsex[ACEs$sexstatus == 1 & ACEs$icsex == 0], 
     add = T, breaks = seq(5,36,2), 
     col = tred)
axis(2, las = 2)
text(x = 30, y = 100, "p < 0.001", cex = 0.9)
mtext("d", at = c(0), cex = 1.3, line = 2) 

# dev.off()
```


### Life history-related outcomes

#### 1) Pairwise correlations

##### Fig. 2
ggcorrplots by icsex
```{r corr}
corr_ACEs <- ACEs

library(ggcorrplot)

## Make age NA for right-censored IC
corr_ACEs$Agefirstconc[corr_ACEs$conc_status == 0 & corr_ACEs$icsex == 0] <- NA
corr_ACEs$agesmoke[corr_ACEs$smokestatus == 0] <- NA
corr_ACEs$agedrink[corr_ACEs$drinkstatus == 0] <- NA
corr_ACEs$ageroman[corr_ACEs$romstatus == 0] <- NA
corr_ACEs$age1stsex[corr_ACEs$sexstatus == 0] <- NA


### FEMALES
table(corr_ACEs$gravidity)

f_ACEs <- corr_ACEs %>%
  filter(icsex == 0) 

## corr
f_matrix <- cor(f_ACEs[ , c("agemens", "Agefirstconc", "gravidity", "agesmoke", "agedrink", "ageroman", "age1stsex")], use = "complete.obs", method = "spearman")
colnames(f_matrix) <- c("Menarche", "First conception", "Gravidity", "First smoke", "First drink", "First romance", "Sexual debut")
rownames(f_matrix) <- c("Menarche", "First conception", "Gravidity", "First smoke", "First drink", "First romance", "Sexual debut")

gg_f2 <- ggcorrplot(f_matrix, type = "upper", lab = T, digits = 2, show.legend = F) +
  theme(plot.margin = margin(8.5,1.5,1,2, "pt"))


### MALES
m_ACEs <- corr_ACEs %>%
  filter(icsex == 1) 

mACEs <- m_ACEs %>%
  mutate(x = 0, y = 0)

m_matrix <- cor(mACEs[ , c("x", "y", "pubicdev", "agesmoke", "agedrink", "ageroman", "age1stsex")], use = "complete.obs", method = "spearman")
colnames(m_matrix) <- c("x", "y", "Tanner score   ", "First smoke", "First drink", "First romance", "Sexual debut")
rownames(m_matrix) <- c("x", "y", "Tanner score   ", "First smoke", "First drink", "First romance", "Sexual debut")
  
gg_m2 <- ggcorrplot(m_matrix, type = "upper", lab = T, digits = 2) +
  theme(plot.margin = margin(9.5,6,64,6.5, "pt"))

## combine plots
ggarrange(plotlist = c(gg_f2, gg_m2), ncol = 2,  labels = "auto", font.label = list(face = "plain"))

# ggsave("Figure2.png", width=20, height=10, units='cm', pointsize = 8, dpi = 1600)

```


### Prep data for models
Create columns for discrete time series data
-- start at 0 (1 year before earliest event), end at event time - X
```{r columns}
range(ACEs$agemens, na.rm = T) # 10
range(ACEs$Agefirstconc, na.rm = T) # 14
range(ACEs$agesmoke, na.rm = T) # 5
range(ACEs$agedrink, na.rm = T) # 5
range(ACEs$ageroman, na.rm = T) # 9
range(ACEs$age1stsex, na.rm = T) # 8

data <- ACEs %>%
   mutate(enter = 0, 
         exit.men = agemens - 9, 
         exit.conc = Agefirstconc - 13, 
         exit.smoke = agesmoke - 4, 
         exit.drink = agedrink - 4,
         exit.rom = ageroman - 8, 
         exit.sex = age1stsex - 7
         )
```

Split sexes
```{r icsex}
f_data <- data %>%
  filter(icsex == 0)

m_data <- data %>%
  filter(icsex == 1)
```


#### Create person period datasets
Discrete time-series analysis: person periods 
```{r personperiods}
## generate person-period datasets 
f_data_men <- dataLong(dataShort = f_data, 
                       timeColumn = "exit.men",
                       eventColumn = "menstatus",
                       timeAsFactor = F) 
f_data_conc <- dataLong(dataShort = f_data, 
                       timeColumn = "exit.conc",
                       eventColumn = "conc_status",
                       timeAsFactor = F) 
f_data_smoke <- dataLong(dataShort = f_data, 
                       timeColumn = "exit.smoke",
                       eventColumn = "smokestatus",
                       timeAsFactor = F) 
f_data_drink <- dataLong(dataShort = f_data, 
                       timeColumn = "exit.drink",
                       eventColumn = "drinkstatus",
                       timeAsFactor = F) 
f_data_rom <- dataLong(dataShort = f_data, 
                       timeColumn = "exit.rom",
                       eventColumn = "romstatus",
                       timeAsFactor = F) 
f_data_sex <- dataLong(dataShort = f_data, 
                       timeColumn = "exit.sex",
                       eventColumn = "sexstatus",
                       timeAsFactor = F) 

m_data_smoke <- dataLong(dataShort = m_data, 
                       timeColumn = "exit.smoke",
                       eventColumn = "smokestatus",
                       timeAsFactor = F) 
m_data_drink <- dataLong(dataShort = m_data, 
                       timeColumn = "exit.drink",
                       eventColumn = "drinkstatus",
                       timeAsFactor = F) 
m_data_rom <- dataLong(dataShort = m_data, 
                       timeColumn = "exit.rom",
                       eventColumn = "romstatus",
                       timeAsFactor = F) 
m_data_sex <- dataLong(dataShort = m_data, 
                       timeColumn = "exit.sex",
                       eventColumn = "sexstatus",
                       timeAsFactor = F) 
```

### Model biodemographic LH traits

#### 1) Females

##### i) Menarche
Baseline hazard -- quadratic fxn

Unadjusted:
```{r unadj men}
## null model
Gompertz_menarche_Baseline_sq <- glm(formula = y ~ timeInt + I(timeInt^2),
                  family = binomial(link = "cloglog"),
                  data = f_data_men)
summary(Gompertz_menarche_Baseline_sq) 


## cumaces model:
Gompertz_menarche_aces <- glm(formula = y ~ timeInt + I(timeInt^2) + cumaces,
                  family = binomial(link = "cloglog"),
                  data = f_data_men)
summary(Gompertz_menarche_aces)
exp(confint(Gompertz_menarche_aces))


## dimensions model:
Gompertz_menarche_thdep <- glm(formula = y ~ timeInt + I(timeInt^2) + n_threat + n_deprivation,
                  family = binomial(link = "cloglog"),
                  data = f_data_men)

summary(Gompertz_menarche_thdep)
exp(confint(Gompertz_menarche_thdep))

```

Adjusted: 
```{r adj men}
## null model
Gompertz_menarche_Baseline_adj <- glm(formula = y ~ timeInt + I(timeInt^2) + scale(assets83) + scale(meduc83) + scale(mommenst),
                  family = binomial(link = "cloglog"),
                  data = f_data_men)
summary(Gompertz_menarche_Baseline_adj) 


## cumaces model:
Gompertz_menarche_aces_adj <- glm(formula = y ~ timeInt + I(timeInt^2) + cumaces + scale(assets83) + scale(meduc83) + scale(mommenst),
                  family = binomial(link = "cloglog"),
                  data = f_data_men)
summary(Gompertz_menarche_aces_adj)
exp(confint(Gompertz_menarche_aces_adj))


## dimensions model:
Gompertz_menarche_thdep_adj <- glm(formula = y ~ timeInt + I(timeInt^2) + n_threat + n_deprivation + scale(assets83) + scale(meduc83) + scale(mommenst),
                  family = binomial(link = "cloglog"),
                  data = f_data_men)

summary(Gompertz_menarche_thdep_adj)
exp(confint(Gompertz_menarche_thdep_adj))

```


##### ii) First conception
Baseline hazard -- quadratic fxn

Unadjusted models:
```{r unadj conc}
### Null model
Gompertz_conc_Baseline_sq <- glm(formula = y ~ timeInt + I(timeInt^2),
                               family = binomial(link = "cloglog"),
                               data = f_data_conc)

summary(Gompertz_conc_Baseline_sq) 


## cumaces model
Gompertz_conc_aces <- glm(formula = y ~ timeInt + I(timeInt^2) + cumaces,
                  family = binomial(link = "cloglog"),
                  data = f_data_conc)
summary(Gompertz_conc_aces)
exp(confint(Gompertz_conc_aces))

## dimensions model
Gompertz_conc_thdep <- glm(formula = y ~ timeInt + I(timeInt^2) + n_threat + n_deprivation,
                  family = binomial(link = "cloglog"),
                  data = f_data_conc)
summary(Gompertz_conc_thdep)
exp(confint(Gompertz_conc_thdep))

```

Adjusted models: 
```{r adj conc}
## null model
Gompertz_conc_Baseline_adj <- glm(formula = y ~ timeInt + I(timeInt^2) + scale(assets83) + scale(meduc83) + scale(mommenst),
                  family = binomial(link = "cloglog"),
                  data = f_data_conc)
summary(Gompertz_conc_Baseline_adj) 


## cumaces model:
Gompertz_conc_aces_adj <- glm(formula = y ~ timeInt + I(timeInt^2) + cumaces + scale(assets83) + scale(meduc83) + scale(mommenst),
                  family = binomial(link = "cloglog"),
                  data = f_data_conc)
summary(Gompertz_conc_aces_adj)
exp(confint(Gompertz_conc_aces_adj))


## dimensions model:
Gompertz_conc_thdep_adj <- glm(formula = y ~ timeInt + I(timeInt^2) + n_threat + n_deprivation + scale(assets83) + scale(meduc83) + scale(mommenst),
                  family = binomial(link = "cloglog"),
                  data = f_data_conc)

summary(Gompertz_conc_thdep_adj)
exp(confint(Gompertz_conc_thdep_adj))

```

##### ii) Gravidity

Unadjusted models:
Zero-inflated neg binomial models
```{r unadj grav}
mean(f_data$gravidity)
var(f_data$gravidity)
var(f_data$gravidity) / mean(f_data$gravidity)

#### null model
grav_null <- glmmTMB(gravidity ~ scale(Age2018intv), data = f_data, family = nbinom2, 
                        ziformula = ~1)
summary(grav_null)
exp(confint(grav_null))

#### cumulative ACEs
grav_cumaces <- glmmTMB(gravidity ~ scale(cumaces) + scale(Age2018intv), data = f_data, family = nbinom2, 
                        ziformula = ~1)
summary(grav_cumaces)
exp(confint(grav_cumaces))

### ACE dimensions
grav_td <- glmmTMB(gravidity ~ scale(n_threat) + scale(n_deprivation) + scale(Age2018intv), data = f_data, family = nbinom2,
                        zi = ~ 1)
summary(grav_td)
exp(confint(grav_td))

```

Adjusted models:
Zero-inflated Poisson models
```{r grav adj}
### Cumulative ACEs
grav_cumaces <- glmmTMB(gravidity ~ scale(cumaces) + scale(assets83) + scale(meduc83) + scale(mommenst) + scale(Age2018intv), data = f_data, family = poisson, 
                        ziformula = ~ 1)
summary(grav_cumaces)
exp(confint(grav_cumaces))


## Subtypes: Threat & Deprivation
grav_td <- glmmTMB(gravidity ~ scale(n_threat) + scale(n_deprivation) + scale(assets83) + scale(meduc83) + scale(mommenst) + scale(Age2018intv), data = f_data, family = poisson, 
                   ziformula = ~1)
summary(grav_td)
exp(confint(grav_td))

```


#### 2) Males

##### i) Advanced pubertal development
Binomial model

Unadjusted models:
```{r unadj mat}
## Make binomial variable for advanced devl (tanner score 4 or 5)
m_data <- m_data %>%
  mutate(pubmat = case_when(
    pubicdev %in% c(4,5) ~ 1,
    pubicdev %in% c(1,2,3) ~ 0
  ))
table(m_data$pubmat)

##null
pub_null <- glm(pubmat ~ scale(icage98), data = m_data, family = "binomial")
summary(pub_null)


## cumulative aces
pub_cumaces <- glm(pubmat ~ scale(cumaces) + scale(icage98), data = m_data, family = "binomial")
summary(pub_cumaces)
exp(confint(pub_cumaces))
exp(coef(pub_cumaces)) 


## threat deprivation
pub_fit2 <- glm(pubmat ~ scale(n_threat) + scale(n_deprivation) + scale(icage98), data = m_data, family = "binomial")
summary(pub_fit2)
exp(confint(pub_fit2))
exp(coef(pub_fit2)) 

```


Adjusted models:
```{r adj mat}
##null
pub_null_adj <- glm(pubmat ~ scale(icage98) + scale(assets83) + scale(meduc83) + scale(mommenst), data = m_data, family = "binomial")
summary(pub_null_adj)


## cum aces
pub_cumaces_adj <- glm(pubmat ~ scale(cumaces) + scale(icage98) + scale(assets83) + scale(meduc83) + scale(mommenst), data = m_data, family = "binomial")
summary(pub_cumaces_adj)
exp(confint(pub_cumaces_adj))
exp(coef(pub_cumaces_adj)) # 1 unit increase in cumaces --> odds of more developed 1.16x larger than the odds of being 1 (no dev)


## threat deprivation
pub_fit2_adj <- glm(pubmat ~ scale(n_threat) + scale(n_deprivation) + scale(icage98) + scale(assets83) + scale(meduc83) + scale(mommenst), data = m_data, family = "binomial")
summary(pub_fit2_adj)
exp(confint(pub_fit2_adj))
exp(coef(pub_fit2_adj)) 

```


#### Fig. 3
K-M plots for female biodemographic traits
```{r menarche and first conc}
tdarkred <- makeTransparent("#661100", 180)
tdgray <- makeTransparent("#888888", 180)


### cumulative ACEs
## Menarche
surv_men1 <- Surv(f_data[f_data$cumaces %in% c(0,3), ]$agemens, f_data[f_data$cumaces %in% c(0,3), ]$menstatus) 
survcurv1 <- survfit(surv_men1 ~ cumaces_factor, data = f_data[f_data$cumaces %in% c(0,3), ]) 
men_cum <- ggsurvplot(survcurv1, 
           palette = c(tdgray, tdarkred), linetype = c(1,1), 
           xlim = c(9,17), break.x.by = 2,
           surv.median.line = "hv",
           xlab = "", ylab = "Menarche", title = "Cumulative ACEs",
           font.x=c(size=12), font.y=c(size=12), font.tickslab = c(size = 12),
           font.title = c(size=12),
           fun = "event", 
           legend.title = "", legend.labs = c("0 ", "3+ "),
           font.legend = c(size=12), 
           ggtheme=custom_theme())
# men_cum + labs(tag = "a")

## Pregnancy
surv_preg1 <- Surv(f_data[f_data$cumaces %in% c(0,3), ]$Agefirstconc, f_data[f_data$cumaces %in% c(0,3), ]$pregstatus) 
survcurv2 <- survfit(surv_preg1 ~ cumaces_factor, data = f_data[f_data$cumaces %in% c(0,3), ]) 
preg_cum <- ggsurvplot(survcurv2, 
           palette = c(tdgray, tdarkred), linetype = c(1,1), 
           xlim = c(14,36), ylim = c(0,1), break.x.by = 5, censor = F,
           surv.median.line = "hv",
           xlab = "Age (years)\n", ylab = "First conception", 
           title = " ", 
           subtitle = " ",
           font.x=c(size=12), font.y=c(size=12), 
           font.tickslab = c(size = 12), font.title = c(size=30), 
           font.subtitle = c(size = 11),
           fun = "event", 
           legend = "none", 
           ggtheme=custom_theme())


### Subtypes
## Menarche - threat
surv_men2 <- Surv(f_data[f_data$n_threat %in% c(0,3), ]$agemens, f_data[f_data$n_threat %in% c(0,3), ]$menstatus) 
survcurvt1 <- survfit(surv_men2 ~ n_threat, data = f_data[f_data$n_threat %in% c(0,3), ]) 
men_t <- ggsurvplot(survcurvt1, 
           palette = c(tdgray, tdarkred), linetype = c(1,1), 
           xlim = c(9,17), break.x.by = 2,
           surv.median.line = "hv",
           xlab = "", ylab = " ", title = "Threat score",
           font.x=c(size=12), font.y=c(size=12), font.tickslab = c(size = 12),
           font.title = c(size=12),
           fun = "event", 
           legend.title = "", legend.labs = c("0 ", "3+ "),
           font.legend = c(size=12), 
           ggtheme=custom_theme())

## First conc - threat
surv_preg2 <- Surv(f_data[f_data$n_threat %in% c(0,3), ]$Agefirstconc, f_data[f_data$n_threat %in% c(0,3), ]$pregstatus) 
survcurvt2 <- survfit(surv_preg2 ~ n_threat, data = f_data[f_data$n_threat %in% c(0,3), ]) 
preg_t <- ggsurvplot(survcurvt2, 
           palette = c(tdgray, tdarkred), linetype = c(1,1), 
           xlim = c(14,36), ylim = c(0,1), break.x.by = 5, censor = F,
           surv.median.line = "hv",
           xlab = "Age (years)\n", ylab = "", 
           title = " ", subtitle = " ",
           font.x=c(size=12), font.y=c(size=12), 
           font.tickslab = c(size = 12), font.title = c(size=30), 
           font.subtitle = c(size = 11),
           fun = "event", 
           legend = "none", 
           ggtheme=custom_theme())


## Menarche - deprivation
surv_men3 <- Surv(f_data[f_data$n_deprivation %in% c(0,3), ]$agemens, f_data[f_data$n_deprivation %in% c(0,3), ]$menstatus) 
survcurvd1 <- survfit(surv_men3 ~ n_deprivation, data = f_data[f_data$n_deprivation %in% c(0,3), ]) 
men_d <- ggsurvplot(survcurvd1, 
           palette = c(tdgray, tdarkred), linetype = c(1,1), 
           xlim = c(9,17), break.x.by = 2,
           surv.median.line = "hv",
           xlab = "", ylab = " ", title = "Deprivation score",
           font.x=c(size=12), font.y=c(size=12), font.tickslab = c(size = 12),
           font.title = c(size=12),
           fun = "event", 
           legend.title = "", legend.labs = c("0 ", "3 "),
           font.legend = c(size=12), 
           ggtheme=custom_theme())

## First conc - deprivation
surv_preg3 <- Surv(f_data[f_data$n_deprivation %in% c(0,3), ]$Agefirstconc, f_data[f_data$n_deprivation %in% c(0,3), ]$pregstatus) 
survcurvd2 <- survfit(surv_preg3 ~ n_deprivation, data = f_data[f_data$n_deprivation %in% c(0,3), ]) 
preg_d <- ggsurvplot(survcurvd2, 
           palette = c(tdgray, tdarkred), linetype = c(1,1), 
           xlim = c(14,36), ylim = c(0,1), break.x.by = 5, censor = F,
           surv.median.line = "hv",
           subtitle = "*",
           xlab = "Age (years)\n", ylab = " ", 
           title = " ", 
           font.x=c(size=12), font.y=c(size=12), 
           font.tickslab = c(size = 12), font.title = c(size=25), 
           font.subtitle = c(size = 16),
           fun = "event", 
           legend = "none", 
           ggtheme=custom_theme())


plot_grid(men_cum$plot,
preg_cum$plot,
men_t$plot,
preg_t$plot,
men_d$plot,
preg_d$plot,
labels = "auto", byrow = F,
label_size = 12,
label_fontface = "plain", 
label_x = c(0.05,0.1,0.1,0.05,0.1,0.1),
label_y = c(0.95,0.95, 0.95,0.92,0.92,0.92))

# ggsave('Figure3.png', width=16, height=13.3, units='cm',  pointsize = 8, dpi = 1600)

```

## Model behavioral LH traits

#### 1) Females

##### i) First smoke 
```{r unadj f smoke}
### Null model
Gompertz_smoke_Baseline_cb <- glm(formula = y ~ timeInt + I(timeInt^2) + I(timeInt^3),
                  family = binomial(link = "cloglog"),
                  data = f_data_smoke)
summary(Gompertz_smoke_Baseline_cb) 

### cumaces model
Gompertz_smoke_aces <- glm(formula = y ~ timeInt + I(timeInt^2) + I(timeInt^3) + cumaces,
                  family = binomial(link = "cloglog"),
                  data = f_data_smoke)
summary(Gompertz_smoke_aces)
exp(confint(Gompertz_smoke_aces))

Gompertz_smoke_td <- glm(formula = y ~ timeInt + I(timeInt^2) + I(timeInt^3) + n_threat + n_deprivation,
                  family = binomial(link = "cloglog"),
                  data = f_data_smoke)
summary(Gompertz_smoke_td)
exp(confint(Gompertz_smoke_td))

```

Adjusted models
```{r adj f smoke}
### Null model
Gompertz_smoke_Baseline_adj <- glm(formula = y ~ timeInt + I(timeInt^2) + I(timeInt^3) + scale(assets83) + scale(meduc83) + scale(mommenst),
                  family = binomial(link = "cloglog"),
                  data = f_data_smoke)
summary(Gompertz_smoke_Baseline_adj) 

### cumaces model
Gompertz_smoke_aces_adj <- glm(formula = y ~ timeInt + I(timeInt^2) + I(timeInt^3) + cumaces + scale(assets83) + scale(meduc83) + scale(mommenst),
                  family = binomial(link = "cloglog"),
                  data = f_data_smoke)
summary(Gompertz_smoke_aces_adj)
exp(confint(Gompertz_smoke_aces_adj))

Gompertz_smoke_td_adj <- glm(formula = y ~ timeInt + I(timeInt^2) + I(timeInt^3) + n_threat + n_deprivation + scale(assets83) + scale(meduc83) + scale(mommenst),
                  family = binomial(link = "cloglog"),
                  data = f_data_smoke)
summary(Gompertz_smoke_td_adj)
exp(confint(Gompertz_smoke_td_adj))

```

##### ii) First drink

Unadjusted models:
```{r unadj f drink}
### Null model
Gompertz_drink_Baseline_cb <- glm(formula = y ~ timeInt + I(timeInt^2) + I(timeInt^3),
                  family = binomial(link = "cloglog"),
                  data = f_data_drink)
summary(Gompertz_drink_Baseline_cb) 


### cumaces model
Gompertz_drink_aces <- glm(formula = y ~ timeInt + I(timeInt^2) + I(timeInt^3) + cumaces,
                  family = binomial(link = "cloglog"),
                  data = f_data_drink)
summary(Gompertz_drink_aces)
exp(confint(Gompertz_drink_aces))

### dimensions model
Gompertz_drink_td <- glm(formula = y ~ timeInt + I(timeInt^2) + I(timeInt^3) + n_threat + n_deprivation,
                               family = binomial(link = "cloglog"),
                               data = f_data_drink)
summary(Gompertz_drink_td)
exp(confint(Gompertz_drink_td))
```

Adjusted models
```{r adj f drink}
### Null model
Gompertz_drink_Baseline_adj <- glm(formula = y ~ timeInt + I(timeInt^2) + I(timeInt^3) + scale(assets83) + scale(meduc83) + scale(mommenst),
                  family = binomial(link = "cloglog"),
                  data = f_data_drink)
summary(Gompertz_drink_Baseline_adj) 


### cumaces model
Gompertz_drink_aces_adj <- glm(formula = y ~ timeInt + I(timeInt^2) + I(timeInt^3) + cumaces + scale(assets83) + scale(meduc83) + scale(mommenst),
                  family = binomial(link = "cloglog"),
                  data = f_data_drink)
summary(Gompertz_drink_aces_adj)
exp(confint(Gompertz_drink_aces_adj))

### dimensions model
Gompertz_drink_td_adj <- glm(formula = y ~ timeInt + I(timeInt^2) + I(timeInt^3) + n_threat + n_deprivation + scale(assets83) + scale(meduc83) + scale(mommenst),
                               family = binomial(link = "cloglog"),
                               data = f_data_drink)
summary(Gompertz_drink_td_adj)
exp(confint(Gompertz_drink_td_adj))

```

##### iii) First romantic relationship

Unadjusted models:
```{r unadj f rom}
### Null model
Gompertz_rom_Baseline_cb <- glm(formula = y ~ timeInt + I(timeInt^2) + I(timeInt^3),
                  family = binomial(link = "cloglog"),
                  data = f_data_rom)
summary(Gompertz_rom_Baseline_cb) 


### cumaces model
Gompertz_rom_aces <- glm(formula = y ~ timeInt + I(timeInt^2) + I(timeInt^3) + cumaces,
                  family = binomial(link = "cloglog"),
                  data = f_data_rom)
summary(Gompertz_rom_aces)
exp(confint(Gompertz_rom_aces))

### dimensions model
Gompertz_rom_td <- glm(formula = y ~ timeInt + I(timeInt^2) + I(timeInt^3) + n_threat + n_deprivation,
                  family = binomial(link = "cloglog"),
                  data = f_data_rom)
summary(Gompertz_rom_td)
exp(confint(Gompertz_rom_td))

```

Adjusted models
```{r romance adj}
### Null model
Gompertz_rom_Baseline_adj <- glm(formula = y ~ timeInt + I(timeInt^2) + I(timeInt^3) + scale(assets83) + scale(meduc83) + scale(mommenst),
                  family = binomial(link = "cloglog"),
                  data = f_data_rom)
summary(Gompertz_rom_Baseline_adj) 


### cumaces model
Gompertz_rom_aces_adj <- glm(formula = y ~ timeInt + I(timeInt^2) + I(timeInt^3) + cumaces + scale(assets83) + scale(meduc83) + scale(mommenst),
                  family = binomial(link = "cloglog"),
                  data = f_data_rom)
summary(Gompertz_rom_aces_adj)
exp(confint(Gompertz_rom_aces_adj))

### dimensions model
Gompertz_rom_td_adj <- glm(formula = y ~ timeInt + I(timeInt^2) + I(timeInt^3) + n_threat + n_deprivation + scale(assets83) + scale(meduc83) + scale(mommenst),
                               family = binomial(link = "cloglog"),
                               data = f_data_rom)
summary(Gompertz_rom_td_adj)
exp(confint(Gompertz_rom_td_adj))

```

##### iv) Sexual debut

Unadjusted models
```{r unadj f sex}
### Null model
Gompertz_sex_Baseline_cb <- glm(formula = y ~ timeInt + I(timeInt^2) + I(timeInt^3),
                  family = binomial(link = "cloglog"),
                  data = f_data_sex)
summary(Gompertz_sex_Baseline_cb) 


### cumaces model
Gompertz_sex_aces <- glm(formula = y ~ timeInt + I(timeInt^2) + I(timeInt^3) + cumaces,
                 family = binomial(link = "cloglog"),
                  data = f_data_sex)
summary(Gompertz_sex_aces)
exp(confint(Gompertz_sex_aces))

### dimensions model
Gompertz_sex_td <- glm(formula = y ~ timeInt + I(timeInt^2) + I(timeInt^3) + n_threat + n_deprivation,
                  family = binomial(link = "cloglog"),
                  data = f_data_sex)
summary(Gompertz_sex_td)
exp(confint(Gompertz_sex_td))

```

Adjusted models
```{r sex adj}
### Null model
Gompertz_sex_Baseline_adj <- glm(formula = y ~ timeInt + I(timeInt^2) + I(timeInt^3) + scale(assets83) + scale(meduc83) + scale(mommenst),
                  family = binomial(link = "cloglog"),
                  data = f_data_sex)
summary(Gompertz_sex_Baseline_adj) 


### cumaces model
Gompertz_sex_aces_adj <- glm(formula = y ~ timeInt + I(timeInt^2) + I(timeInt^3) + cumaces + scale(assets83) + scale(meduc83) + scale(mommenst),
                  family = binomial(link = "cloglog"),
                  data = f_data_sex)
summary(Gompertz_sex_aces_adj)
exp(confint(Gompertz_sex_aces_adj))


### dimensions model
Gompertz_sex_td_adj <- glm(formula = y ~ timeInt + I(timeInt^2) + I(timeInt^3) + n_threat + n_deprivation + scale(assets83) + scale(meduc83) + scale(mommenst),
                               family = binomial(link = "cloglog"),
                               data = f_data_sex)
summary(Gompertz_sex_td_adj)
exp(confint(Gompertz_sex_td_adj))

```


#### 2) Males

##### i) First smoke 
```{r unadj m smoke}
### Null model
mGompertz_smoke_Baseline_cb <- glm(formula = y ~ timeInt + I(timeInt^2) + I(timeInt^3),
                  family = binomial(link = "cloglog"),
                  data = m_data_smoke)
summary(mGompertz_smoke_Baseline_cb) 

### cumaces model
mGompertz_smoke_aces <- glm(formula = y ~ timeInt + I(timeInt^2) + I(timeInt^3) + cumaces,
                  family = binomial(link = "cloglog"),
                  data = m_data_smoke)
summary(mGompertz_smoke_aces)
exp(confint(mGompertz_smoke_aces))

mGompertz_smoke_td <- glm(formula = y ~ timeInt + I(timeInt^2) + I(timeInt^3) + n_threat + n_deprivation,
                  family = binomial(link = "cloglog"),
                  data = m_data_smoke)
summary(mGompertz_smoke_td)
exp(confint(mGompertz_smoke_td))

```

Adjusted models
```{r adj m smoke}
### Null model
mGompertz_smoke_Baseline_adj <- glm(formula = y ~ timeInt + I(timeInt^2) + I(timeInt^3) + scale(assets83) + scale(meduc83) + scale(mommenst),
                  family = binomial(link = "cloglog"),
                  data = m_data_smoke)
summary(mGompertz_smoke_Baseline_adj) 

### cumaces model
mGompertz_smoke_aces_adj <- glm(formula = y ~ timeInt + I(timeInt^2) + I(timeInt^3) + cumaces + scale(assets83) + scale(meduc83) + scale(mommenst),
                  family = binomial(link = "cloglog"),
                  data = m_data_smoke)
summary(mGompertz_smoke_aces_adj)
exp(confint(mGompertz_smoke_aces_adj))

mGompertz_smoke_td_adj <- glm(formula = y ~ timeInt + I(timeInt^2) + I(timeInt^3) + n_threat + n_deprivation + scale(assets83) + scale(meduc83) + scale(mommenst),
                  family = binomial(link = "cloglog"),
                  data = m_data_smoke)
summary(mGompertz_smoke_td_adj)
exp(confint(mGompertz_smoke_td_adj))

```

##### ii) First drink

Unadjusted models:
```{r unadj m drink}
### Null model
mGompertz_drink_Baseline_cb <- glm(formula = y ~ timeInt + I(timeInt^2) + I(timeInt^3),
                  family = binomial(link = "cloglog"),
                  data = m_data_drink)
summary(mGompertz_drink_Baseline_cb) 


### cumaces model
mGompertz_drink_aces <- glm(formula = y ~ timeInt + I(timeInt^2) + I(timeInt^3) + cumaces,
                  family = binomial(link = "cloglog"),
                  data = m_data_drink)
summary(mGompertz_drink_aces)
exp(confint(mGompertz_drink_aces))

### dimensions model
mGompertz_drink_td <- glm(formula = y ~ timeInt + I(timeInt^2) + I(timeInt^3) + n_threat + n_deprivation,
                               family = binomial(link = "cloglog"),
                               data = m_data_drink)
summary(mGompertz_drink_td)
exp(confint(mGompertz_drink_td))
```

Adjusted models
```{r adj m drink}
### Null model
mGompertz_drink_Baseline_adj <- glm(formula = y ~ timeInt + I(timeInt^2) + I(timeInt^3) + scale(assets83) + scale(meduc83) + scale(mommenst),
                  family = binomial(link = "cloglog"),
                  data = m_data_drink)
summary(mGompertz_drink_Baseline_adj) 


### cumaces model
mGompertz_drink_aces_adj <- glm(formula = y ~ timeInt + I(timeInt^2) + I(timeInt^3) + cumaces + scale(assets83) + scale(meduc83) + scale(mommenst),
                  family = binomial(link = "cloglog"),
                  data = m_data_drink)
summary(mGompertz_drink_aces_adj)
exp(confint(mGompertz_drink_aces_adj))

### dimensions model
mGompertz_drink_td_adj <- glm(formula = y ~ timeInt + I(timeInt^2) + I(timeInt^3) + n_threat + n_deprivation + scale(assets83) + scale(meduc83) + scale(mommenst),
                               family = binomial(link = "cloglog"),
                               data = m_data_drink)
summary(mGompertz_drink_td_adj)
exp(confint(mGompertz_drink_td_adj))

```

##### iii) First romantic relationship

Unadjusted models:
```{r unadj m rom}
### Null model
mGompertz_rom_Baseline_cb <- glm(formula = y ~ timeInt + I(timeInt^2) + I(timeInt^3),
                  family = binomial(link = "cloglog"),
                  data = m_data_rom)
summary(mGompertz_rom_Baseline_cb) 


### cumaces model
mGompertz_rom_aces <- glm(formula = y ~ timeInt + I(timeInt^2) + I(timeInt^3) + cumaces,
                  family = binomial(link = "cloglog"),
                  data = m_data_rom)
summary(mGompertz_rom_aces)
exp(confint(mGompertz_rom_aces))

### dimensions model
mGompertz_rom_td <- glm(formula = y ~ timeInt + I(timeInt^2) + I(timeInt^3) + n_threat + n_deprivation,
                  family = binomial(link = "cloglog"),
                  data = m_data_rom)
summary(mGompertz_rom_td)
exp(confint(mGompertz_rom_td))

```

Adjusted models
```{r m romance adj}
### Null model
mGompertz_rom_Baseline_adj <- glm(formula = y ~ timeInt + I(timeInt^2) + I(timeInt^3) + scale(assets83) + scale(meduc83) + scale(mommenst),
                  family = binomial(link = "cloglog"),
                  data = m_data_rom)
summary(mGompertz_rom_Baseline_adj) 


### cumaces model
mGompertz_rom_aces_adj <- glm(formula = y ~ timeInt + I(timeInt^2) + I(timeInt^3) + cumaces + scale(assets83) + scale(meduc83) + scale(mommenst),
                  family = binomial(link = "cloglog"),
                  data = m_data_rom)
summary(mGompertz_rom_aces_adj)
exp(confint(mGompertz_rom_aces_adj))

### dimensions model
mGompertz_rom_td_adj <- glm(formula = y ~ timeInt + I(timeInt^2) + I(timeInt^3) + n_threat + n_deprivation + scale(assets83) + scale(meduc83) + scale(mommenst),
                               family = binomial(link = "cloglog"),
                               data = m_data_rom)
summary(mGompertz_rom_td_adj)
exp(confint(mGompertz_rom_td_adj))

```

##### iv) Sexual debut

Unadjusted models
```{r unadj m sex}
### Null model
mGompertz_sex_Baseline_cb <- glm(formula = y ~ timeInt + I(timeInt^2) + I(timeInt^3),
                  family = binomial(link = "cloglog"),
                  data = m_data_sex)
summary(mGompertz_sex_Baseline_cb) 


### cumaces model
mGompertz_sex_aces <- glm(formula = y ~ timeInt + I(timeInt^2) + I(timeInt^3) + cumaces,
                 family = binomial(link = "cloglog"),
                  data = m_data_sex)
summary(mGompertz_sex_aces)
exp(confint(mGompertz_sex_aces))

### dimensions model
mGompertz_sex_td <- glm(formula = y ~ timeInt + I(timeInt^2) + I(timeInt^3) + n_threat + n_deprivation,
                  family = binomial(link = "cloglog"),
                  data = m_data_sex)
summary(mGompertz_sex_td)
exp(confint(mGompertz_sex_td))

```

Adjusted models
```{r m sex adj}
### Null model
mGompertz_sex_Baseline_adj <- glm(formula = y ~ timeInt + I(timeInt^2) + I(timeInt^3) + scale(assets83) + scale(meduc83) + scale(mommenst),
                  family = binomial(link = "cloglog"),
                  data = m_data_sex)
summary(mGompertz_sex_Baseline_adj) 


### cumaces model
mGompertz_sex_aces_adj <- glm(formula = y ~ timeInt + I(timeInt^2) + I(timeInt^3) + cumaces + scale(assets83) + scale(meduc83) + scale(mommenst),
                  family = binomial(link = "cloglog"),
                  data = m_data_sex)
summary(mGompertz_sex_aces_adj)
exp(confint(mGompertz_sex_aces_adj))


### dimensions model
mGompertz_sex_td_adj <- glm(formula = y ~ timeInt + I(timeInt^2) + I(timeInt^3) + n_threat + n_deprivation + scale(assets83) + scale(meduc83) + scale(mommenst),
                               family = binomial(link = "cloglog"),
                               data = m_data_sex)
summary(mGompertz_sex_td_adj)
exp(confint(mGompertz_sex_td_adj))

```

#### Fig. 4

Make function to plot survival curves together
```{r plot fxn}
genCombSurvPlots <- function(modellist, data, outcome, xlab, ylab, minage, maxage) {
  plt <- ggsurvplot_combine(modellist, data = data, 
          # palette = c("#888888D2", "#661100D2", tdgray, tdarkred),
           palette = c(alpha("#888888", 0.7), alpha("#661100", 0.7), alpha("#888888", 0.8), alpha("#661100", 0.8)),
           linetype = c(1,1,3,3), 
           break.x.by = 5, ylim = c(0,1),
           surv.median.line = "hv",
           xlim = c(minage, maxage),
           xlab = xlab, ylab = ylab,
           title = outcome, font.title = c(size=12),
           subtitle = " ", font.subtitle = c(size = 11),
           font.x=c(size=12), font.y=c(size=12), font.tickslab = c(size = 12),
           fun = "event", censor = F, legend = "none",
           # legend.title = "", legend.labs = c("0 F", "3+ F", "0 M", "3+ M"),
           # font.legend = c(size=12),
           ggtheme=custom_theme())
  return(plt)
}

## function to grab legend
genLegend <- function(modellist, data, outcome, xlab, ylab, minage, maxage) {
  plt <- ggsurvplot_combine(modellist, data = data, 
          # palette = c("#888888D2", "#661100D2", tdgray, tdarkred),
           palette = c(alpha("#888888", 0.7), alpha("#661100", 0.7), alpha("#888888", 0.8), alpha("#661100", 0.8)),
           linetype = c(1,1,3,3), 
           break.x.by = 5, ylim = c(0,1),
           surv.median.line = "hv",
           xlim = c(minage, maxage),
           xlab = xlab, ylab = ylab,
           title = outcome, font.title = c(size=14),
           subtitle = " ", font.subtitle = c(size = 11),
           font.x=c(size=12), font.y=c(size=12), font.tickslab = c(size = 12),
           fun = "event", censor = F, 
          # legend.position = "right",
           legend.title = "Cumulative\nACEs", legend.labs = c("0 F", "3+ F", "0 M", "3+ M"),
           font.legend = c(size=10),
           ggtheme=custom_theme())
           #  theme(
           #   legend.key.spacing.y = unit(5, "pt")
           # )) 
  return(plt)
  }
```

Combine survival curves (male and female)
```{r beh sexes comb}
tdarkred <- makeTransparent("#661100", 180)
tdgray <- makeTransparent("#888888", 180)

## SMOKING 
#### F:
surv_smokeF <- Surv(f_data[f_data$cumaces %in% c(0,3), ]$agesmoke, f_data[f_data$cumaces %in% c(0,3), ]$smokestatus) 
survcurv1 <- survfit(surv_smokeF ~ cumaces_factor, data = f_data[f_data$cumaces %in% c(0,3), ]) 

#### M:
surv_smokeM <- Surv(m_data[m_data$cumaces %in% c(0,3), ]$agesmoke, m_data[m_data$cumaces %in% c(0,3), ]$smokestatus) 
survcurv2 <- survfit(surv_smokeM ~ cumaces_factor, data = m_data[m_data$cumaces %in% c(0,3), ]) 

smoke_surv_models <- list(survcurv1, survcurv2)


## DRINKING 
#### F:
surv_drinkF <- Surv(f_data[f_data$cumaces %in% c(0,3), ]$agedrink, f_data[f_data$cumaces %in% c(0,3), ]$drinkstatus) 
survcurv1 <- survfit(surv_drinkF ~ cumaces_factor, data = f_data[f_data$cumaces %in% c(0,3), ]) 

#### M:
surv_drinkM <- Surv(m_data[m_data$cumaces %in% c(0,3), ]$agedrink, m_data[m_data$cumaces %in% c(0,3), ]$drinkstatus) 
survcurv2 <- survfit(surv_drinkM ~ cumaces_factor, data = m_data[m_data$cumaces %in% c(0,3), ]) 

drink_surv_models <- list(survcurv1, survcurv2)

## ROMANCE 
#### F:
surv_romF <- Surv(f_data[f_data$cumaces %in% c(0,3), ]$ageroman, f_data[f_data$cumaces %in% c(0,3), ]$romstatus) 
survcurv1 <- survfit(surv_romF ~ cumaces_factor, data = f_data[f_data$cumaces %in% c(0,3), ]) 

#### M:
surv_romM <- Surv(m_data[m_data$cumaces %in% c(0,3), ]$ageroman, m_data[m_data$cumaces %in% c(0,3), ]$romstatus) 
survcurv2 <- survfit(surv_romM ~ cumaces_factor, data = m_data[m_data$cumaces %in% c(0,3), ]) 

rom_surv_models <- list(survcurv1, survcurv2)


## SEX 
#### F:
surv_sexF <- Surv(f_data[f_data$cumaces %in% c(0,3), ]$age1stsex, f_data[f_data$cumaces %in% c(0,3), ]$sexstatus) 
survcurv1 <- survfit(surv_sexF ~ cumaces_factor, data = f_data[f_data$cumaces %in% c(0,3), ]) 

#### M:
surv_sexM <- Surv(m_data[m_data$cumaces %in% c(0,3), ]$age1stsex, m_data[m_data$cumaces %in% c(0,3), ]$sexstatus) 
survcurv2 <- survfit(surv_sexM ~ cumaces_factor, data = m_data[m_data$cumaces %in% c(0,3), ]) 

sex_surv_models <- list(survcurv1, survcurv2)


### get legend for multipanel figure
legendplot <- genLegend(smoke_surv_models, data, "First smoke", xlab = " ", ylab = "Probability", minage = 10, maxage = 25)

legendplot2 <- legendplot$plot + guides(color = guide_legend(ncol = 1, direction = "vertical"))

# extract the legend from one of the plots
legend <- ggpubr::get_legend(
  legendplot2 + theme(legend.box.margin = margin(0, 0, 0, 18))
)

## or horizontal legend
legendplot_h <- legendplot$plot + guides(color = guide_legend(ncol = 2))

# extract the legend from one of the plots
legend_h <- ggpubr::get_legend(
  legendplot_h + theme(legend.box.margin = margin(0, 0, 10, 0))
)
```

Generate plots
```{r fig 4 plots}
smokePlt <- genCombSurvPlots(smoke_surv_models, data, "First smoke", xlab = " ", ylab = "Probability", minage = 10, maxage = 25)
drinkPlt <- genCombSurvPlots(drink_surv_models, data, outcome = "First drink", xlab = " ", ylab = " ", minage = 10, maxage = 25)
romPlt <- genCombSurvPlots(rom_surv_models, data, "First relationship", xlab = "Age (years)", ylab = "Probability", minage = 10, maxage = 25)
sexPlt <- genCombSurvPlots(sex_surv_models, data, "Sexual debut", xlab = "Age (years)", ylab = " ", minage = 15, maxage = 30)
```

Combine multipanel figure
```{r fig 4}
prows <- plot_grid(smokePlt$plot,
          drinkPlt$plot,
          romPlt$plot,
          sexPlt$plot, 
labels = "auto", byrow = T,
label_size = 12,
label_fontface = "plain",
label_x = c(0.05))

prows

## combine with legend (on side)
plot_grid(prows, legend, nrow = 1, ncol = 2, rel_widths = c(2,0.4))

# ggsave('Figure4.pdf', width=16, height=13.3, units='cm',  pointsize = 8, dpi = 1600)

```